---
title: "Hurricane data analyis"
author: "Achim Rumberger"
date: "27 Januar 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

## Analysis of Hurricane data from HURDAT package


```{r}
#libraries
library(tidyverse)
library(ggthemes)
library(ggmap)
library(htmlwidgets)
library(gridExtra)
library(HURDAT)
library(lubridate)

source("classifiy_hurricianes.R")

```


## Read Inital and Save Data
The data span the time from 1851 to 2016.
We will have almost 80k observations with 24 variables.
It will take a while to read the inital data with get_hurdat function. 
Therefore I will save the data for more easy access in the future.
If you run this code the first time please uncomment the lines with "get_hurdat" and "saveRDS".
```{r}
#data <- get_hurdat(basin = c("AL", "EP"))
#saveRDS(data, "data/hurricanesALEP")
data <- readRDS("data/hurricanesALEP")

```

##Manage the dates
```{r}
data$YEAR <- year(data$DateTime)
data$MONTH <- month(data$DateTime)
data$DAY <- day(data$DateTime)
summary(data %>% select(YEAR, MONTH, DAY, Wind, Pressure))
```


## First plot to get an overview of the data
```{r}
# first plot
df = data %>%
  group_by(YEAR) %>%
  summarise(Distinct_Storms = n_distinct(Key))

p = ggplot(df, aes(x = YEAR, y = Distinct_Storms)) + theme_stata()
p + geom_line(size = 1.1) + 
  ggtitle("Number of Storms Per Year") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_smooth(method='lm', se = FALSE) + 
  ylab("Storms")
```

##Categorize storms
Create a new variable which classifies the hurricanes according to windspeed.

```{r}
data$CATEGORY <- sapply(data$Wind, classifyHurricane)

```

## Plot the storms by category
```{r fig.height=20, fig.width=12}
#by category
df = data %>%
  filter(grepl("H", CATEGORY)) %>%
  group_by(YEAR,CATEGORY) %>%
  summarise(Distinct_Storms = n_distinct(Key))
df$CATEGORY = factor(df$CATEGORY)

p = ggplot(df, aes(x = YEAR, y = Distinct_Storms, col = CATEGORY)) + theme_stata()
p + geom_line(size = 1.1) + 
  scale_color_brewer(direction = -1, palette = "Dark2") + 
  ggtitle("Number of Storms Per Year By Category (H)") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ CATEGORY, scales = "free_x", ncol = 1) + 
  geom_smooth(method = 'lm', se = FALSE, col = 'black') +
  theme(axis.text.x = element_text(hjust = 1, angle=45), legend.position = 'none') + 
  ylab('Number of Storms') +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)))
```
## Make a distinct dataset
I also will filter the data. A hurricane in its lifecycle will go through several stages, from tropical storm to a class H1 hurricane to H2 to H1, tropical storm. I want to count each storm only once with its highest ratings.
On my laptop this takes a while, so again, I will save the data for further analysis. This will show us, that in the time frame from 1851 to 2016 a total of 2900 storms have been recorded.
If you run this code the first time please uncomment the lines with "highestClassification4Storm" and "saveRDS".

```{r}
#distinctStormData <- highestClassification4Storm(data)
#saveRDS(distinctStormData, "data/hurricanesALEPAAA")
distinctStormData <- readRDS("data/hurricanesALEPAAA")
summary(distinctStormData)
```
## Plot all distinct storms

```{r}
df = distinctStormData %>%
  group_by(YEAR) %>%
  summarise(Distinct_Storms = n_distinct(Key))

p = ggplot(df, aes(x = YEAR, y = Distinct_Storms)) + theme_stata()
p + geom_line(size = 1.1) + 
  ggtitle("Number of distinct Storms Per Year") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_smooth(method='lm', se = FALSE) + 
  ylab('Number of Storms') +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)))

```
## Plot disitinct hurricanes by category

```{r fig.height=20, fig.width=12}
df = distinctStormData %>%
  filter(grepl("H", CATEGORY)) %>%
  group_by(YEAR,CATEGORY) %>%
  summarise(Distinct_Storms = n_distinct(Key))
df$CATEGORY = factor(df$CATEGORY)

p = ggplot(df, aes(x = YEAR, y = Distinct_Storms, col = CATEGORY)) + theme_stata()
p + geom_line(size = 1.1) + 
  scale_color_brewer(direction = -1, palette = "Dark2") + 
  ggtitle("Number ofd disitinct Storms Per Year By Category (H)") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~CATEGORY, scales = "free_x", ncol = 1) + 
  geom_smooth(method = 'lm', se = FALSE, col = 'black') +
  theme(axis.text.x = element_text(angle=45), legend.position = 'none') + 
  ylab('Number of Storms') +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)))
```